---
title: "R Assessment 2"
output: html_notebook
    number_sections: true
    toc: true
    toc_depth: 4
    code_folding: hide
    theme: cosmo
    highlight: tango
---

# Executive Summary 

This compiled dataset pulled from four other datasets linked by time and place, and was built to find signals correlated to increased suicide rates among different cohorts globally, across the socio-economic spectrum.

# Installing important libraries
```{r}
# deleted them for the knitting
```
# Loading important packages
```{r}
library(knitr)
library(dplyr)#cleaning
library(readr)
library(ggplot2)#to create graphs
library(tibble)#to create the graphs
library(stringr)
library(gridExtra)
library(scales)
library(lubridate)
library(ggrepel)
library(leaflet)
library(highcharter) # Interactive data visualizations
library(plotly) # Interactive data visualizations
library(viridis) # Color gradients
library(tidyverse) # Metapackge
library(countrycode) # Converts country names/codes
library(rjson) # JSON reader
library(crosstalk) # Provides interactivity for HTML widgets
library(DT) # Displaying data tables
library(shiny)# to create shiny files
library(shinyWidgets)
```
#read in data
* 2016 data was removed (few countries had any, those that did often had data missing)
```{r}
data <- read.csv('/Users/hasimrahman/Documents/master.csv') %>%
filter(year != 2016, # filter out 2016 and countries with 0 data. 
         country != 'Dominica',
         country != 'Saint Kitts and Nevis')

```

Next we check the data 

glimpse function gives us the data structure and the data size
```{r}
glimpse(data)
```
* 7 countries removed (<= 3 years of data total)
* HDI was removed due to 2/3 missing data
* Africa has very few countries providing suicide data

# Fix the names of some of the countries in our data to match the country names 
# used by our map later on so that they'll be interpreted and displayed. 

```{r}
data <- data %>%
  mutate(country = fct_recode(country, "The Bahamas" = "Bahamas"),
         country = fct_recode(country, "Cape Verde" = "Cabo Verde"),
         country = fct_recode(country, "South Korea" = "Republic of Korea"),
         country = fct_recode(country, "Russia" = "Russian Federation"),
         country = fct_recode(country, "Republic of Serbia" = "Serbia"),
         country = fct_recode(country, "United States of America" = "United States"))
```
# Reorder levels of age to be in chronological order.
```{r}
data$age <- factor(data$age, levels = c("5-14 years", "15-24 years", "25-34 years", "35-54 years", "55-74 years", "75+ years"))
```
Check for missing data
```{r}
# Check for missing values in the entire dataset
missing_values <- colSums(is.na(data))

# Display columns with missing values and their counts
print(missing_values[missing_values > 0])

```
Removing the missing value by removing the column

```{r}
# Check for missing values in the HDI.for.year column
missing_values_hdi <- sum(is.na(data$HDI.for.year))

# Display the number of missing values
print(paste("Missing values in HDI.for.year:", missing_values_hdi))

# Remove the HDI.for.year column
data <- subset(data, select = -c(HDI.for.year))

```


Create a theme for the charts
```{r}
# Create a custom theme for the plots with blue and orange colors.
custom_theme <- hc_theme(
  colors = c('#5CACEE', 'orange'),  # Change colors to blue and orange
  chart = list(
    backgroundColor = '#FAFAFA',
    plotBorderColor = "black"
  ),
  xAxis = list(
    gridLineColor = "E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    title = list(style = list(color = "#333333"))
  ),
  yAxis = list(
    gridLineColor = "#E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    tickWidth = 1,
    title = list(style = list(color = "#333333"))
  ),
  title = list(style = list(color = '#333333', fontFamily = "Lato")),
  subtitle = list(style = list(color = '#666666', fontFamily = "Lato")),
  legend = list(
    itemStyle = list(color = "#333333"),
    itemHoverStyle = list(color = "#FFF"),
    itemHiddenStyle = list(color = "#606063")
  ),
  credits = list(style = list(color = "#666")),
  itemHoverStyle = list(color = 'gray')
)

```

# Graph 1 
We are creating a line graph to show different levles of suicide in throughout the years, this interactive time series graph gives clear trend and insight on how the suicide data changed over the years


define the user interface

```{r}
ui <- fluidPage(
  titlePanel("Worldwide Suicides"),
  sidebarLayout(
    sidebarPanel(
      h4("Explanation"),
      p("This interactive application allows you to explore worldwide suicides over the years."),
      p("Use the slider to select a specific year range, and the line graph will show the trends in suicides."),
      p("The black dashed line represents the mean suicides per 100K people."),
      p("Explore and analyze the data to gain insights."),
      hr(),
      h4("Variables"),
      p("1. Year: The year of the record."),
      p("2. Suicides_no: Count of suicides."),
      p("3. Population: Population of the country."),
      p("4. Suicide_capita: Suicides per 100K population (calculated)."),
      hr(),
      p("Select Year Range:"),
      sliderInput("year_slider", "Year Range", 
                  min = min(data$year), max = max(data$year),
                  value = c(min(data$year), max(data$year)), step = 1)
    ),
    mainPanel(
      highchartOutput("line_plot")
    )
  )
)

# Your server function remains unchanged

server <- function(input, output) {
  # Create tibble for our line plot.
  overall_tibble <- reactive({
    data %>%
      filter(year >= input$year_slider[1], year <= input$year_slider[2]) %>%
      select(year, suicides_no, population) %>%
      group_by(year) %>%
      summarise(suicide_capita = round((sum(suicides_no) / sum(population)) * 100000, 2))
  })

  # Render the line plot using highcharter with your custom theme.
  output$line_plot <- renderHighchart({
    highchart() %>%
      hc_add_series(overall_tibble(), hcaes(x = year, y = suicide_capita, color = suicide_capita), type = "line") %>%
      hc_tooltip(crosshairs = TRUE, borderWidth = 1.5, headerFormat = "", 
                 pointFormat = paste("Year: <b>{point.x}</b> <br> Suicides: <b>{point.y}</b>")) %>%
      hc_title(text = "Worldwide suicides by year") %>%
      hc_subtitle(text = paste("Selected Years: ", input$year_slider[1], "-", input$year_slider[2])) %>%
      hc_xAxis(title = list(text = "Year")) %>%
      hc_yAxis(title = list(text = "Suicides per 100K people"),
               allowDecimals = FALSE,
               plotLines = list(
                 list(
                   color = "black", width = 1, dashStyle = "Dash",
                   value = mean(overall_tibble()$suicide_capita),
                   label = list(text = "Mean = 13.12",
                                style = list(color = "black", fontSize = 11))
                 )
               )) %>%
      hc_legend(enabled = FALSE) %>%
      hc_add_theme(custom_theme)
  })
}
shinyApp(ui = ui, server = server)
```

# Graph 2
Next we create a world graph to see how different countries were effected by suicide. This bar plot divides all the  countires to their age groups.
In this graph we have an option to choose which countries we want to compare, this bar plot shows the suicide per 100k people for every country, the colour further denotees the amount of deaths, we also have a slider to choose the year and a box option to choose the countries.

```{r}
custom_theme <- hc_theme(
  colors = c('#5CACEE', 'orange'),  # Change colors to blue and orange
  chart = list(
    backgroundColor = '#FAFAFA',
    plotBorderColor = "black"
  ),
    xAxis = list(
    gridLineColor = "E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    title = list(style = list(color = "#333333"))
  ),
  yAxis = list(
    gridLineColor = "#E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    tickWidth = 1,
    title = list(style = list(color = "#333333"))
  ),
  title = list(style = list(color = '#333333', fontFamily = "Lato")),
  subtitle = list(style = list(color = '#666666', fontFamily = "Lato")),
  legend = list(
    itemStyle = list(color = "#333333"),
    itemHoverStyle = list(color = "#FFF"),
    itemHiddenStyle = list(color = "#606063")
  ),
  credits = list(style = list(color = "#666")),
  itemHoverStyle = list(color = 'gray')
)
```

creating the rshiny
```{r}
ui <- fluidPage(
  titlePanel("Worldwide Suicides"),
  sidebarLayout(
    sidebarPanel(
      h4("Explanation"),
      p("This interactive application allows you to explore worldwide suicides over the years."),
      p("Use the slider to select a specific year range, and the bar plot will show the suicides per 100K people for selected countries."),
      hr(),
      h4("Variables"),
      p("1. Year: The year of the record."),
      p("2. Suicides_no: Count of suicides."),
      p("3. Population: Population of the country."),
      p("4. Country: The countries."),
      hr(),
      p("Select Year Range:"),
      sliderInput("year_slider", "Year Range", 
                  min = min(data$year), max = max(data$year),
                  value = c(min(data$year), max(data$year)), step = 1),
      hr(),
      selectInput("countries", "Select Countries", choices = unique(data$country), multiple = TRUE)
    ),
    mainPanel(
      highchartOutput("bar_plot")
    )
  )
)

server <- function(input, output) {
  # Create tibble for overall suicides by selected countries
  selected_countries_bar <- reactive({
    data %>%
      filter(country %in% input$countries, year >= input$year_slider[1], year <= input$year_slider[2]) %>%
      group_by(country) %>%
      summarise(suicide_capita = round((sum(suicides_no) / sum(population)) * 100000, 2)) %>%
      arrange(desc(suicide_capita))
  })

  output$bar_plot <- renderHighchart({
    highchart() %>%
      hc_add_series(selected_countries_bar(), hcaes(x = country, y = suicide_capita, color = suicide_capita), type = "bar") %>%
      hc_tooltip(borderWidth = 1.5, 
                 pointFormat = paste("Suicides: <b>{point.y}</b>")) %>%
      hc_legend(enabled = FALSE) %>%
      hc_title(text = "Suicides by selected countries") %>%
      hc_subtitle(text = "1985-2015") %>%
      hc_xAxis(categories = selected_countries_bar()$country,
               labels = list(step = 1),
               min = 0, max = 25,
               scrollbar = list(enabled = TRUE)) %>%
      hc_yAxis(title = list(text = "Suicides per 100K people")) %>%
      hc_plotOptions(bar = list(stacking = "normal", 
                                pointPadding = 0, groupPadding = 0, borderWidth = 0.5)) %>%
      hc_add_theme(custom_theme)
  })
}

shinyApp(ui = ui, server = server)

```


# graph 3
This graph shows the world map and we cang hover through and see the countries and their suicide population, we also have a slider to choose the years, with this we can study particular years or huge range of it and it can easily answer our question.

```{r}
library(shiny)
library(highcharter)
library(dplyr)


# Create a custom theme for the plots.
custom_theme <- hc_theme(
  colors = c('#5CACEE', 'orange'),  # Change colors to blue and orange
  chart = list(
    backgroundColor = '#FAFAFA',
    plotBorderColor = "black"
  ),
  xAxis = list(
    gridLineColor = "E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    title = list(style = list(color = "#333333"))
  ),
  yAxis = list(
    gridLineColor = "#E5E5E5",
    labels = list(style = list(color = "#333333")),
    lineColor = "#E5E5E5",
    minorGridLineColor = "#E5E5E5",
    tickColor = "#E5E5E5",
    tickWidth = 1,
    title = list(style = list(color = "#333333"))
  ),
  title = list(style = list(color = '#333333', fontFamily = "Lato")),
  subtitle = list(style = list(color = '#666666', fontFamily = "Lato")),
  legend = list(
    itemStyle = list(color = "#333333"),
    itemHoverStyle = list(color = "#FFF"),
    itemHiddenStyle = list(color = "#606063")
  ),
  credits = list(style = list(color = "#666")),
  itemHoverStyle = list(color = 'gray')
  
)

# UI definition
ui <- fluidPage(
  titlePanel("Suicides by Country"),
  sidebarLayout(
    sidebarPanel(
      h4("Explanation"),
      p("This interactive application allows you to explore worldwide suicides over the years."),
      p("The world map shows suicides per 100K people by country for the specified year range."),
      hr(),
      h4("Variables"),
      p("1. Year: The year of the record."),
      p("2. Suicides_no: Count of suicides."),
      p("3. Population: Population of the country."),
      p("4. Suicide_capita: Suicides per 100K population (calculated)."),
      hr(),
      sliderInput("year_range", "Select Year Range", 
                  min = min(data$year), max = max(data$year),
                  value = c(min(data$year), max(data$year)), step = 1)
    ),
    mainPanel(
      highchartOutput("world_map")
    )
  )
)

# Server logic
server <- function(input, output) {
  # Create a tibble with suicide per capita by country for the specified year range.
  country_tibble <- reactive({
    data %>%
      filter(year >= input$year_range[1], year <= input$year_range[2]) %>%
      group_by(country) %>%
      summarise(suicide_capita = round((sum(suicides_no) / sum(population)) * 100000, 2))
  })

  # Render the interactive world map
  output$world_map <- renderHighchart({
    highchart() %>%
      hc_add_series_map(worldgeojson, country_tibble(), value = "suicide_capita", joinBy = c('name', 'country')) %>%
      hc_colorAxis(stops = color_stops()) %>%
      hc_title(text = "Suicides by Country") %>%
      hc_subtitle(text = paste("Selected Years: ", input$year_range[1], "-", input$year_range[2])) %>%
      hc_tooltip(borderWidth = 1.5, headerFormat = "", valueSuffix = " suicides (per 100K people)") %>%
      hc_add_theme(custom_theme)
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```
#Graph 4
In this interactive option for the graph we are able to compare all of the variables. We have different values to put into the scatter plot

1. X Axis: Horizontal axis displaying the variable of interest values.
2. Y Axis: Vertical axis indicating the measure or dependent variable.
3. Measure Color: Hue representing a third dimension for additional data distinction.
4. Facet Row: Rows splitting data based on a categorical variable.
5. Facet Column: Columns dividing data for detailed categorical visualization.
With these we can add the variables to these to create many types of charts so the user can gather better data from it.

```{r}
library(shiny)
library(ggplot2)


# UI definition
ui <- fluidPage(
  headerPanel("Data Visualization of Suicide Dataset"),
  pageWithSidebar(
    headerPanel("Suicide Dataset Analysis"),
    sidebarPanel(
      h4('The variables are as follows:'),
      h6('1.  country               : Country name'),
      h6('2.  year                  : Year of the record'),
      h6('3.  sex                   : Gender (male/female)'),
      h6('4.  age                   : Age group'),
      h6('5.  suicides_no           : Count of suicides'),
      h6('6.  population            : Population of the country'),
      h6('7.  suicide_rate          : Suicides per 100K population'),
      h6('8.  country-year          : Composite key of country and year'),
      h6('9.  HDI for year          : Human Development Index for the year'),
      h6('10. gdp_for_year ($)      : GDP for the year in dollars'),
      h6('11. gdp_per_capita ($)    : GDP per capita in dollars'),
      h6('12. generation            : Generation based on age grouping average'),
      br(),
      h4('Select different parameters:'),
      sliderInput('sampleSize', 'Sample Size', min = 1, max = nrow(data),
                  value = min(5000, nrow(data)), step = 100, round = 0),
      selectInput('x', 'X Axis Measure', names(data)),
      selectInput('y', 'Y Axis Measure', names(data), names(data)[[6]]),
      selectInput('color', 'Measure Color', c('None', names(data)), names(data)[[4]]),
      selectInput('facet_row', 'Facet Row', c(None = '.', names(data)), names(data)[[2]]),
      selectInput('facet_col', 'Facet Column', c(None = '.', names(data))),
      textInput('caption', 'Plot Caption', value = 'Plot on Suicide Dataset')
    ),
    mainPanel(
      h4('Introduction & How to use'),
      p("A suicide dataset contains information about suicides worldwide with various attributes. This Shiny application allows you to interactively explore the dataset by selecting different parameters for plotting."),
      br(),
      plotOutput('plot')
    )
  )
)

# Server logic
server <- function(input, output) {
  # Reload the data based on Sample Size selected by the user
  dataset <- reactive({
    data[sample(nrow(data), input$sampleSize), ]
  })
  
  # Plot based on user inputs
  output$plot <- renderPlot({
    p <- ggplot(dataset(), aes_string(x = input$x, y = input$y)) + geom_point(aes_string(color = input$color))
    
    facets <- paste(input$facet_row, '~', input$facet_col)
    if (facets != '. ~ .') {
      p <- p + facet_grid(facets)
    }
    
    p <- p + ggtitle(input$caption) +
      scale_color_brewer(palette = "Set3")  # You can change the color palette here
    
    print(p)
  }, height = 800)
}

shinyApp(ui = ui, server = server)

```

```{r}
library(shiny)
library(ggplot2)

# Assuming 'data' is your suicide dataset
# Replace 'data' with your actual dataset name if different

# UI definition
ui <- fluidPage(
  headerPanel("Data Visualization of Suicide Dataset"),
  pageWithSidebar(
    headerPanel("Suicide Dataset Analysis"),
    sidebarPanel(
      h4('The variables are as follows:'),
      h6('1.  country               : Country name'),
      h6('2.  year                  : Year of the record'),
      h6('3.  sex                   : Gender (male/female)'),
      h6('4.  age                   : Age group'),
      h6('5.  suicides_no           : Count of suicides'),
      h6('6.  population            : Population of the country'),
      h6('7.  suicide_rate          : Suicides per 100K population'),
      h6('8.  country-year          : Composite key of country and year'),
      h6('9.  HDI for year          : Human Development Index for the year'),
      h6('10. gdp_for_year ($)      : GDP for the year in dollars'),
      h6('11. gdp_per_capita ($)    : GDP per capita in dollars'),
      h6('12. generation            : Generation based on age grouping average'),
      br(),
      h4('Select different parameters:'),
      sliderInput('sampleSize', 'Sample Size', min = 1, max = nrow(data),
                  value = min(5000, nrow(data)), step = 100, round = 0),
      selectInput('x', 'X Axis Measure', names(data)),
      selectInput('color', 'Measure Color', c('None', names(data)), names(data)[[4]]),
      textInput('caption', 'Plot Caption', value = 'Density Plot on Suicide Dataset'),
      selectInput('plot_type', 'Choose Visualization Type',
                  choices = c('Density Plot', 'Bar Plot'), selected = 'Density Plot')
    ),
    mainPanel(
      h4('Introduction & How to use'),
      p("A suicide dataset contains information about suicides worldwide with various attributes. This Shiny application allows you to interactively explore the dataset by selecting different parameters and visualization types for plotting."),
      br(),
      plotOutput('plot')
    )
  )
)

# Server logic
server <- function(input, output) {
  # Reload the data based on Sample Size selected by the user
  dataset <- reactive({
    data[sample(nrow(data), input$sampleSize), ]
  })
  
  # Plot based on user inputs and selected visualization type
  output$plot <- renderPlot({
    p <- switch(input$plot_type,
                'Density Plot' = ggplot(dataset(), aes_string(x = input$x, fill = input$color)) +
                  geom_density(alpha = 0.5) +
                  ggtitle(input$caption) +
                  scale_fill_brewer(palette = "Set3"),
                
                'Bar Plot' = ggplot(dataset(), aes_string(x = input$x, y = ..count.., fill = input$color)) +
                  geom_bar(stat = "count", position = "dodge") +
                  ggtitle(input$caption) +
                  scale_fill_brewer(palette = "Set3")
    )
    
    print(p)
  }, height = 800)
}

shinyApp(ui = ui, server = server)



```
#Graph 5
In this interactive graph we are able to make bar plots and density plots of all the variables
Through this graph the user is able to make different graphs and study every variable with different representations the changeable slider options are.
1. Change Countries:Modify the displayed data by selecting specific countries of interest.
2. X Axis: Horizontal axis, representing the primary variable in the visual.
3. Measure Color: Hue indicating an additional dimension for data differentiation.
4. Choose Visualization Type: Select the preferred graphical representation for data exploration.

```{r}
library(shiny)
library(ggplot2)

# Assuming 'data' is your suicide dataset
# Replace 'data' with your actual dataset name if different

# UI definition
ui <- fluidPage(
  headerPanel("Data Visualization of Suicide Dataset"),
  pageWithSidebar(
    headerPanel("Suicide Dataset Analysis"),
    sidebarPanel(
      h4('The variables are as follows:'),
      h6('1.  country               : Country name'),
      h6('2.  year                  : Year of the record'),
      h6('3.  sex                   : Gender (male/female)'),
      h6('4.  age                   : Age group'),
      h6('5.  suicides_no           : Count of suicides'),
      h6('6.  population            : Population of the country'),
      h6('7.  suicide_rate          : Suicides per 100K population'),
      h6('8.  country-year          : Composite key of country and year'),
      h6('9.  HDI for year          : Human Development Index for the year'),
      h6('10. gdp_for_year ($)      : GDP for the year in dollars'),
      h6('11. gdp_per_capita ($)    : GDP per capita in dollars'),
      h6('12. generation            : Generation based on age grouping average'),
      br(),
      h4('Select different parameters:'),
      sliderInput('sampleSize', 'Sample Size', min = 1, max = nrow(data),
                  value = min(5000, nrow(data)), step = 100, round = 0),
      selectInput('countries', 'Choose Countries',
                  choices = unique(data$country), multiple = TRUE),
      selectInput('x', 'X Axis Measure', names(data)),
      selectInput('color', 'Measure Color', c('None', names(data)), names(data)[[4]]),
      textInput('caption', 'Plot Caption', value = 'Visualization Plot on Suicide Dataset'),
      selectInput('plot_type', 'Choose Visualization Type',
                  choices = c('Density Plot', 'Histogram', 'Scatter Plot'), selected = 'Density Plot')
    ),
    mainPanel(
      h4('Introduction & How to use'),
      p("A suicide dataset contains information about suicides worldwide with various attributes. This Shiny application allows you to interactively explore the dataset by selecting different parameters, visualization types, and specific countries for plotting."),
      br(),
      plotOutput('plot')
    )
  )
)

# Server logic
server <- function(input, output) {
  # Reload the data based on Sample Size and selected countries
  dataset <- reactive({
    data[data$country %in% input$countries, ][sample(nrow(data), input$sampleSize), ]
  })
  
  # Plot based on user inputs and selected visualization type
  output$plot <- renderPlot({
    p <- switch(input$plot_type,
                'Density Plot' = ggplot(dataset(), aes_string(x = input$x, fill = input$color)) +
                  geom_density(alpha = 0.5) +
                  ggtitle(input$caption) +
                  scale_fill_brewer(palette = "Set3"),
                
                'Histogram' = ggplot(dataset(), aes_string(x = input$x, fill = input$color)) +
                  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
                  ggtitle(input$caption) +
                  scale_fill_brewer(palette = "Set3"),
                
                'Scatter Plot' = ggplot(dataset(), aes_string(x = input$x, y = population, color = input$color)) +
                  geom_point(alpha = 0.5) +
                  ggtitle(input$caption) +
                  scale_fill_brewer(palette = "Set3")
    )
    
    print(p)
  }, height = 800)
}

shinyApp(ui = ui, server = server)


```

